---
title: "GeorefRham"
author: "Laura Kostyniuk"
date: "04/08/2020"
output: html_document
---

# **Rhamnus cathartica: Georeferencing and Making Species Distribution Models**

This project's goal is to convert the spreadsheet data of georeferenced locations of instances of R. cathartica into maps projections and models that can overlay climate data, or separate the data into different subcategories.

R. cathartica (common buckthorn) is native to Europe, parts of Asia, and parts of Africa. It is naturalized in North America, and considered an invasive species in some US states and Ontario, Canada. 

##GBIF Data Cleansing

The original data that was used in these models is from GBIF. The following steps were applied to the spreadsheets downloaded from GBIF to organize and complete the data:
- re-organize records by country code, province, and locality alphabetically.  
- search and remove existing latitude/longitude coordinates that did not make sense (0,0 coordinates, coordinates that were duplicated in multiple records with different localities).  
- Add missing province information to records that contained clear locality information and/or coordinates already.  
- Manually pass through data: reading, interpreting, and finding the localities described. Once found, the coordinates and uncertainty were obtained using Google Maps and inputted into the spreadsheet.  
- *NOTE: not all localities were possible to identify. In these cases, no coordinates or uncertainty were added to the spreadsheet, and LocationNotFound keyword was added to the record.*  

##Data formatting and Model Construction

The cleansed data from GBIF was downloaded as a .tsv file and formatted properly for model construction in R studio. The following steps were applied to the data in this Rmd file.
- Download necessary R packages before they will be needed.  
- Import data from Google sheets.  
- Refine data format for use in R studio.  
- Create a local copy of refined dataset.  
- Plot data on scatterplot to see general shape (checking that data makes sense).  
- Plot data on a basic map projection.  
- Create subsets of data; by country, decade, etc.  
- Colour code points on map by subsets created.  

- Create climate data maps  
- Add uncertainty to points  
```{r}
# packages required to install to run this Rmd file
install.packages(c("googlesheets4", 
                   "raster", 
                   "rgdal", 
                   "dismo", 
                   "rJava", 
                   "maptools", 
                   "rgeos", 
                   "scales"))
```
Reading Google Sheets data and saving a local copy with odd columns removed:
```{r}
library(googlesheets4) # loading package

Rham <- read_sheet("https://docs.google.com/spreadsheets/d/1JCgvTShYHXt2PfiDQowLeX9m3N7h_2r8mqTk6vS6_tU/edit?usp=sharing") # read data
#must copy and paste above line into terminal
```

```{r}
Rham <- Rham[, ! colnames(Rham) %in% c("occurrenceID", "georeferencedDate",
                                       "catalogNumber", "recordNumber")] # removing funky columns
write.table(x = Rham, file = "Rham.tsv") # storing local copy of file from google sheets to prevent having to re-authenticate
Rham <- read.table("Rham.tsv") # data preview
```
Visualizing when the occurrences of R. cathartica are noted during the year (by month):
```{r}
hist(Rham$month, breaks = 12, xlab="Month")
```
Comparing the worldwide frequency of occurrence per month to the Canadian occurrence per month:
```{r}
hist(cad$month, breaks = 12, xlab="Month")
```
Histrogram of distribution of years most data was collected:
```{r}
hist(Rham$year, breaks = 420, xlab="Decade")
```
```{r}
hist(log(Rham$coordinateUncertaintyInMeters), breaks=50)
hist(Rham$coordinateUncertaintyInMeters, breaks=50)
```
Plotting out all the data we will be working with on a simple XY scatterplot:
```{r}
plot(x = Rham$decimalLongitude, y = Rham$decimalLatitude, xlab = "Longitude", ylab = "Latitude", cex=0.5)

```
Plotting the data we will be working with on a map projection:
```{r}
library(scales)
library(rgeos)
library(maptools)
data(wrld_simpl)
plot(wrld_simpl)
box()
points(Rham$decimalLongitude, Rham$decimalLatitude, col=alpha("red", 0.5), pch=16, cex=0.25)
```
A summary of all the information about the dataset:
```{r}
summary(Rham) # important information: min year = 1600, max year = 2019
              # important information: min precision = 1, max precision = 4000000
```

```{r}
# creating objects; practice/useful?
withDate <- subset(Rham, year >= 1600 & year <= 2019)
withPrec <- subset(Rham, coordinateUncertaintyInMeters >= 1 & coordinateUncertaintyInMeters <= 4000000)
```
The following map is the same world map zoomed in on the US. The colour of the dot corresponds to the decade the record was recorded.
```{r}
usa <- subset(Rham, countryCode == "US")
usa$decade <- cut(usa$year, seq(1829, 2019, by = 10), dig.lab=4)

library(scales)
library(rgeos)
library(maptools)
data(wrld_simpl)

plot(wrld_simpl, xlim=c(-120, -63), ylim=c(10, 50))
box()
points(x=usa$decimalLongitude, y=usa$decimalLatitude, col=alpha(colours()[usa$decade]), pch=16, cex=0.25)
legend(x=-140, y=50, legend=levels(usa$decade), col=(colours()[usa$decade]), pch=20, cex = 0.5)
```
The following map is the same world map zoomed in on Canada. The colour of the dot corresponds to the decade the record was recorded.
```{r}
cad <- subset(Rham, countryCode == "CA")
cad$decade <- cut(cad$year, seq(1879, 2019, by = 10), dig.lab=4)

library(scales)
library(rgeos)
library(maptools)
data(wrld_simpl)

plot(wrld_simpl, xlim=c(-145, -50), ylim=c(40, 80))
box()
#levels(cad$decade) <- c(seq(cad$decade))
points(x=cad$decimalLongitude, y=cad$decimalLatitude, col=alpha(colours()[cad$decade], 0.75), cex=0.25, pch=16)
legend(x=-170, y=80, legend=levels(cad$decade), col=colours()[cad$decade], pch=20, cex = 0.75)
```
More work would have to be done to adjust the colours and make them more intuitive. I would like to be able to use hex codes for colours and create smooth gradients. such is possible with [this website](http://www.perbang.dk/rgbgradient/)

The following is the setup of the North American data to make a more informative map. 
```{r}
#setup of Noth American data
nam <- bind(cad, usa)
print("The earliest record collected in North America:", quote=FALSE)
min(nam$year, na.rm=TRUE)
print("The youngest record collected in North America:", quote=FALSE)
max(nam$year, na.rm=TRUE)

nam$decade <- cut(nam$year, seq(1839, 2019, by = 10), dig.lab=4)

nam$quell <- cut(nam$year, seq(1824, 2024, by = 25), dig.lab=4)
#all year values are one less than desired for cutoff rates as cut function creates (] levels; therefore and exclusive beginning and inclusive ending number, and the values given define the endings of each level

#adding "No Year" levels to both decade and quell columns
print("Table of decade levels", quote=FALSE)
levels(nam$decade) <- c(levels(nam$decade), "No Year")
nam[is.na(nam$year), "decade"] <- "No Year"
table(nam$decade)

print("Table of quarter century levels", quote=FALSE)
levels(nam$quell) <- c(levels(nam$quell), "No Year")
nam[is.na(nam$year), "quell"] <- "No Year"
table(nam$quell)
```
The following is a setup of the world data. It is the same setup as the North American data, however we will use the North American data for now to create a map of distirbution in North America.
```{r}
#setup of world data
print("The earliest record collected:", quote=FALSE)
min(Rham$year, na.rm=TRUE)
print("The youngest record collected:", quote=FALSE)
max(Rham$year, na.rm=TRUE)
Rham$decade <- cut(Rham$year, seq(1599, 2019, by = 10), dig.lab=4)

Rham$quell <- cut(Rham$year, seq(1599, 2024, by = 25), dig.lab=4)
#all year values are one less than desired for cutoff rates as cut function creates (] levels; therefore and exclusive beginning and inclusive ending number, and the values given define the endings of each level

#adding "No Year" levels to both decade and quell columns
print("Table of decade levels", quote=FALSE)
levels(Rham$decade) <- c(levels(Rham$decade), "No Year")
Rham[is.na(Rham$year), "decade"] <- "No Year"
table(Rham$decade)

print("Table of quarter century levels", quote=FALSE)
levels(Rham$quell) <- c(levels(Rham$quell), "No Year")
Rham[is.na(Rham$year), "quell"] <- "No Year"
table(Rham$quell)
```

```{r}
min(nam$coordinateUncertaintyInMeters, na.rm=TRUE)
max(nam$coordinateUncertaintyInMeters, na.rm=TRUE)
nam$uncertainty <- nam$coordinateUncertaintyInMeters
nam[is.na(nam$uncertainty), "uncertainty"] <- 111000
table(nam$uncertainty)
```
The following map is a map of North America. The Canada and USA maps that were bound together are more precise map polygons. On this map, the uncertainty of the point determines and is approximately proportional to its size. The colour corresponds to the decade the record was recorded in.
```{r}
library(raster) #setup of map
library(scales)
canmap <- getData("GADM", country="CAN",level=1)
usamap <- getData("GADM", country="US", level=1)
nammap <-bind(canmap,usamap)
plot(nammap, xlim=c(-170,-60)) #plotting map and points
points(x=nam$decimalLongitude, y=nam$decimalLatitude, col=alpha(colours()[nam$decade], 0.25), cex=nam$uncertainty/111000, pch=16)
#size of points is the uncertainty in meters divided by the approximate meters in a latitude/longitude degree, in an attempt to make it proportional to the map
```
Map of Ontario records, using a grayscale to indicate quarter century:
```{r}
ont <- subset(Rham, Rham$stateProvince == "Ontario")
ont$decColours <- as.integer(ont$quell)-min(as.integer(ont$quell))
ont$decColours <- as.integer(ont$decColours)/max(as.integer(ont$decColours))
ont$decColours <- gray(ont$decColours)
library(raster)
library(scales)
plot(canmap, xlim=c(-90,-73), ylim=c(40,57))
box()
points(x=ont$decimalLongitude, y=ont$decimalLatitude, pch=16, col=ont$decColours, cex=0.5)
legend(x=-105, y=57, legend=levels(ont$quell), col=ont$decColours, pch=20, cex=0.75)
```

Map of North America with uncertainty and quarter centuries:
```{r}
library(raster) #setup of map
library(scales)
canmap <- getData("GADM", country="CAN",level=1)
usamap <- getData("GADM", country="US", level=1)
nammap <-bind(canmap,usamap)
plot(nammap, xlim=c(-170,-60)) #plotting map and points
points(x=nam$decimalLongitude, y=nam$decimalLatitude, col=alpha(colours()[nam$quell], 0.25), cex=nam$coordinateUncertaintyInMeters/111000, pch=16)
#size of points is the uncertainty in meters divided by the approximate meters in a latitude/longitude degree, in an attempt to make it proportional to the map
legend(x=-160, y=55, legend=levels(nam$quell), col=(colours()[nam$quell]), cex=0.6, pch=20)
```
the same map with standard point size:
```{r}
library(raster) #setup of map
library(scales)
canmap <- getData("GADM", country="CAN",level=1)
usamap <- getData("GADM", country="US", level=1)
nammap <-bind(canmap,usamap)
plot(nammap, xlim=c(-170,-60)) #plotting map and points
points(x=nam$decimalLongitude, y=nam$decimalLatitude, col=alpha(colours()[nam$quell], 0.25), cex=0.5, pch=16)
#size of points is the uncertainty in meters divided by the approximate meters in a latitude/longitude degree, in an attempt to make it proportional to the map
legend(x=-160, y=55, legend=levels(nam$quell), col=(colours()[nam$quell]), cex=0.6, pch=20)
```
An attempt to create a map of north america with a colour gradient to represent the year:
```{r}
#install.packages("wesanderson")
library(wesanderson)
library(raster)
library(ggplot2)
#wes_palette("Zissou1")
#pal<-wes_palette("Zissou1", 100, type="continuous")

colpal <- colorRampPalette(c("red", "orange", "yellow", "green", "light blue", "purple"))
nam$col <- colpal(100)[as.numeric(cut(nam$year, breaks=100))]
plot(nam$decimalLongitude, nam$decimalLatitude, pch=16, cex=0.5, col=nam$col, xlim=c(-140,-60))
plot(nammap, xlim=c(-170,-60), add=TRUE)
#legend(x=-140, y=55, legend=nam$year, col=scale_fill_gradientn(colours=colpal))
legend_image <- as.raster(matrix(colpal(100), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'Years')
text(x=1.5, y = seq(0,1,l=5), labels = seq(1840,2020,l=5))
rasterImage(legend_image, 0, 0, 1,1)
```
Creating heat maps:
```{r}
# remove records with missing coordinates
library(raster)
RhamComplete <- subset(Rham, !is.na(decimalLongitude) &
                            !is.na(decimalLatitude)) 

# convert data.frame to a spatialPoints object
coordinates(RhamComplete)  <- ~decimalLongitude+decimalLatitude

# create an empty raster for the whole world
r <- raster(wrld_simpl)

# set to 3 degrees resolution
res(r) = 3

occDensity <- rasterize(coordinates(RhamComplete), r, fun = 'count',
                       background = 0)

plot(occDensity, xlim = c(-100, -60), ylim = c(35, 50))
plot(occDensity)
plot(wrld_simpl, add = TRUE)

#llproj <- CRS('+proj=longlat +datum=WGS84')
```
A North American heat map:
```{r}
library(raster)
a <- raster(bind(getData("GADM", country="CAN",level=1), getData("GADM", country="US",level=1)))
res(a)=1
occDensity <- rasterize(coordinates(RhamComplete), r, fun = 'count',
                       background = 0)
plot(occDensity, xlim = c(-140, -50), ylim = c(25, 75))
#plot(occDensity)
plot(bind(getData("GADM", country="CAN",level=1), getData("GADM", country="US",level=1)), add = TRUE)
```
Beginning to work with world climate data:
```{r}
library(raster)
wc <- getData(name = "worldclim", var = "bio", res = 5)

plot(wc$bio1)
hist(wc$bio1)
```

```{r}
RhamBio1 <- extract(wc$bio1, Rham[ , c("decimalLongitude", "decimalLatitude")])
hist(RhamBio1)
rb1D <- density (RhamBio1, na.rm=TRUE)
wcb1D <- density(wc$bio1, na.rm=TRUE, plot=FALSE)
ylims <-c(0, max(rb1D$y, wcb1D$y))
plot(wcb1D, col="red", ylim=ylims, main="Bio1")
lines(density(RhamBio1, na.rm=TRUE))

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
